Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists

### 2024-11-06 paper.Rmd
Corrections (Karsten/Kenneth)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
# read data
pre_surveys <- readRDS("data/pre_workshop_surveys.rds")
post_surveys <- readRDS("data/post_workshop_surveys.rds")
# factore levels for charts
factor_levels <- c("Excellent", "Good", "Sufficient", "Insufficient", "Not shared")
# readable event names
event_names <- as_labeller(c("202304_CAA" = "CAA", "202308_EAA" = "EAA", "202310_CAA-DE-NLFl" = "CAA-DE-NLFl", "202311_Reuvensdagen" = "Reuvensdagen", "202311_CAA_UK" = "CAA_UK", "202312_Leiden" = "Leiden", "202401_Aarhus" = "Aarhus"))

@@

Computational modelling has become a standard research technique across most scientific disciplines, including archaeology. Among the many computational modelling techniques the discipline has found agent-based modelling to be particularly well suited to the questions asked by archaeologists. Agent-based modelling is a simulation technique where autonomous agents interact with each other and with their environment. Their aggregated behaviour gives rise to system-level outcomes thus enabling researchers to understand how *micro* (individual) *mechanisms* lead to *macro* (system-level) *patterns*.

While Agent-Based Modelling importance in archaeological practice is steadily growing [@romanowska2023] there are significant bariers to its wider adoption. One of the sectorial problems is the access to vocational training required to gain the necessary computer programming and modelling skills. A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 76% of modellers were primarily self-taught with some degree of secondary support i.e. 27% relied on peer support, only 7% had any sort of formal training at a workshop or summer school, while only 14% received training as part of a degree.

A consortium of Landward Research Teoranta, Aarhus Universitet, Universiteit Leiden and Saxion University of Applied Sciences recognised this need, and sought to address the limited opportunities that archaeologists have to access high quality training through a project aiming to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.

@@

-   development of a hands-on vocational training programme, aligned with the recently published *Agent-Based Modelling for Archaeology: Simulating the Complexity of Societies* [@romanowska2021a];

@@

3.  improved opportunities for archaeologists to work in other European countries (transnational mobility) through matching the skills being delivered with the European Commission's Digital Competence Framework for Citizens (2.2) and the sectoral Skills Passport for Archaeology, which will mean that it will be easier for individual workers to demonstrate their skills to employers across Europe;

@@

The ABMA learning materials assume that participants have learning skills at least at a secondary education level. It is suitable for participants with at least some background in archaeology, such as those typical among students after the first year of a Bachelor study in Archaeology, i.e., above EQF levels 4 or 5. In addition, the tutorials are aimed at professionals working in archaeology who have some experience with computer applications. We also assume that learners are proficient in English and have at least the Reading B1 level, but Reading B2 is recommended [@councilofeurope2020]. The tutorials were also developed with a secondary aim to develop learners' general skills in the following competence area's: 1. Information and data literacy, 2. Communication and collaboration, 3. Digital content creation and, 5. Problem solving [@europeancommissionjointresearchcentre2022].

After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events all of which were run by members of the project team. Recognising the difficulty of delivering practical training in an online environment during the online events several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things. Throughout the project we sought feedback from the participants and used it to improve the tutorials prior to the following events. The feedback was obtained through two surveys using Qualtrics (<https://www.qualtrics.com/>). The participants of the events were asked to respond to two surveys: one before the start of the event and one completed after the end of the event. The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix 1 for the questions). At the end of the event the participants were asked to answer a second survey. Questions of this survey were aimed at measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix 2 for the questions). For some events participants had to register beforehand and we were able to send the pre-workshop survey by email. At other events no registration was possible or necessary and the pre-workshop survey was given at the start of the event. The post-workshop survey was distributed using QR-codes or links at the end of the event.

| Conference or event          | Period                  | Online, in person or hybrid | Number of participants | Number of registrations |
|---------------|---------------|---------------|---------------|---------------|
| CAA Amsterdam                | April 2023              | In person                   | 36                     | 53                      |
| EAA Belfast                  | August 2023             | In person                   | 21                     | No registration         |
| CAA-DE/NL-Fl Online workshop | October 2023            | Online                      | 40                     | 76                      |
| Reuvensdagen                 | November 2023           | In person                   | 13                     | No registration         |
| CAA-UK                       | November 2023           | Hybrid                      | 53                     | No registration         |
| Leiden (course)              | December 2023           | In person                   | 11                     | No registration         |
| Aarhus                       | January/February 2024   | Online                      | 176                    | \>500                   |
| Saxion                       | March 2023-January 2024 | In person                   | 18                     | No registration         |
| **Total**                    |                         |                             | **368**                | **\>628**               |

: Conferences, events and situations were workshops were held at which the tutorials were tested.

@@

The final set of ABMA tutorals are available for download at <https://github.com/ABMArchaeologists/ABMA_tutorials> and are citeable [@rocks-macqueen2024]. The set of developed tutorials consists of the following tutorials:

@@

Each tutorials consists of a number of individual lessons, each of which guides the learner in a self-pace manner through an increasingly challenging set of exercises that build the skills and knowledge necessary to become a confident agent-based modeller.

In the first tutorial the user learn what simulations and agent-based models are and how they can help in archaeological research (\@ref(fig:screenshot-lesson-3)). This tutorial consists of 4 lessons. The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific. Various concepts related to ABM, such as the definitions of model and simulation, are introduced and the participants make their first steps in NetLogo syntax. The third lesson explains how ABMs are used in archaeological research. The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode. This tutorial builds a foundation for working in NetLogo and with ABM. All activity stays within the entree level of difficulty, although some higher level concepts are lightly touched upon.

@@

In tutorial 2 the user will learn the basics of NetLogo by making your their simulation on the Out of Africal dispersal of *Homo sapiens* [@young1995].The learner works through the basics of NetLogo syntax and learn how to set up a simulation and visualize its outcomes. This tutorial consists of 9 lessons. This tutorial intends to guide the learner from the introduction to the beginner level of proficiency. The learning curve is relatively gentle. This is achieved by working with the NetLogo web interface that enables highlighting specific words and buttons and giving visual cues as to how to proceed. The tutorial intoduces several most commonly used built-in functions in Netlogo (called *primitives*) are introduces the general structure of all models cosisting of the initialization phase and the main simulation loop. It walks the participants through te most important NetLogo settings, such as dimensions, coordinates and the origin point of the grid on which the simulation runs. The learners gain their first stripes in programming by using simple loops and functions as well as conditional statements. The next level is achieved by introducing the topic of function definition and different types of variables. Exporting data produced by the model through plots is also explained.

In the third tutorial, the learner builds a simple trade model, again based on a published simulation [@romanowska2018]. Here learners are expanding their NetLogo programming skills with more complexity added to such syntactic structures as loops, lists and reporters. The learner is introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity. This tutorial consists of 7 lessons. This tutorial intends to guide the learner from the beginner to the intermediate level of proficiency. This is achieved by introducing standards of code development such as modular code, pseudocode, debugging and annotating the code. In addition, the tutorial introduces custom agent breeds, visualization with labels and reporters, plots and monitors. This takes the learner to a more advanced level, since it develops knowledge of the programming language combined with diagnosing unexpected behaviour and problem solving.

In tutorial 4 the learner works with a seminal Sugarscape model [@epstein1996] to further expand their agent-based modelling and NetLogo skills. The learner explores how to set up more complex interactions between agents and the environment. Furthermore, the principles of setting up good experiments and validating models is explained. This tutorial consists of 8 lessons. This tutorial intends to solidify the learner's understanding at the intermediate level of proficiency by introducing yet more complex agent-environment interaction, alternative ways to visualize the environment and modelling spatial dynamics. Creating toy landscapes is central to lesson 5. The learner dives into the topic of experiment design at a practical level but also with regards to the scientific principles, such as validation. Thus they are guided through the process of setting up experiments in NetLogo and collecting results using monitors and plots. The topic of validation of agent-based models is critical for most archaelogical applications and thus several ways to compare simulation results with the archaeological record are explored.

In the final and fifth tutorial the participant learn smore about how to incorporate agent-based modelling in archaeological research. This tutorial focuses less on programming in NetLogo and more on the principle and standards of computer-based modelling. The learner learns about the different phases of the model development process. This tutorial consists of 5 lessons. This tutorial summarizes the material delivered in the previous lessons into one coherent framework consistent with an intermediate level of proficiency. This tutorial approaches more theoretical aspects in a practical environment. The learner revisits all phases of the model development process starting with the conceptual phase and finishing with the analysis and interpretation of results. There is a particular focus on discussing what kind of research questions are suitable for modelling and how to pick the right modelling technique. The importance of properly conceptualizing a model before starting the technical phase of the model development is highlighted as it is a common issue for less experienced modellers. The experiment design phase, including parametrisation, validation and the analysis and interpretation of the models is explained at length since it often prove challenging for students. On a practical side, the BehaviorSpace - NetLogo's experiment environment - is explained to enable the learners to batch run their models. Finally the dissemination phase of the model development is also included to ensure that the standards in model publication are widely known. Finally, the learners explores the fundamental importance of replication for advancing scientific knowledge.

@@

A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants). The respondents came from at least 40 countries that represented almost all continents (see Figure \@ref(fig:nationality). There was a clear skew towards participants from Europe. The number of female respondents slightly outnumbers the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \@ref(fig:gender-age)). The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years old indicating that the audience was predominantly composed of early career archaeologists. It seems that the different events also had a slightly different distribution of both gender and age. For example, more older people attended the workshop at the CAA conference in April 2023 and more men were present during the workshop at the Reuvensdagen in November 2023. This might be due the the differences of audiences at the conferences. For the Reuvensdagen the number of participants of the workshop was relatively low and this variance might be due to chance.

@@

```{r gender-age, fig.cap="The gender and age distribution of the respondents for each workshop", fig.width=11, echo=FALSE}
ggplot(pre_surveys, aes(x = `What.is.your.age.`, fill = `What.is.your.gender....Selected.Choice`)) +
  geom_bar() + labs(x= "Age", y="Count", fill = "Gender") + facet_grid(~Workshop, labeller = event_names) +   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

@@

```{r, available-theory, fig.cap="Respondents optionion on the quality of theory on ABM faceted out by event.", fig.width=11, echo=FALSE}
pre_surveys %>%
  select(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`, Workshop) %>%
  filter(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`!="") %>%
  mutate(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.` =
           fct_relevel(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`,
                       "No opinion", "Limited", "Sufficient", "More than sufficient", "Very good")) %>%
  ggplot(aes(x = `How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`)) +
  geom_bar() + xlab("Quality of theory on ABM") + coord_flip() +
  facet_grid(~Workshop, labeller = event_names)
```

@@

During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials \@ref(fig:workshop-impression). Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general. Individual progress varied depending on skills and preferences of the participants. While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later installments.

@@

of the respondents were enthusiastic about the teaching material, with the majority rating it as excellent or good. The teachers were rated even higher than the teaching material, with more respondents giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).

The majority of the respondents (153) were enthusiastic about the teaching material, with the majority rating it as excellent or good. The teachers were even rated better than the teaching material, with more respondents (165) giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).

@@

The respondents generally liked the interactive step-by-step way of going through the tutorials, but also valued the interaction with the teachers strongly. The easy and intuitive introduction into ABM and NetLOGO were also mentioned quite often. This was also observed during the workshops as most participants were working in the own pace and did not need much help. From a didactic point of view, the self-guided and self-paced way of going through the tutorials is very efficient for teachers and limits the frustration of students that may otherwise go too fast or fall behind the class when the tempo is dictated by the teacher.

We also received feedback on possible improvements. During the first workshop, the CAA-conference in Amsterdam in April 2023, we did the first test of the tutorials and we also instructed the participants that this was work in progress. For this and the following few workshops we mostly received feedback about bugs causing the tutorial to crash or creating unexpected behaviour. The number of references to bugs reduced to zero by the end of the project and participants started providing other feedback, for example the wish to see more examples or expanding the tutorials further. Some wanted to introduce more cooperation between participants, which we did not consider primarily aiming the tutorials for individual work. A more cooperative aspect of the education materials is possible, even if harder for online events, and we hope that a wider classroom use of the ABMA materials may deliver solutions. It was also interesting to note that over time more and more people did not see any additional room for improvement indicating that the level, content and format of the tutorials met their needs.

We also asked the participants about their intentions in terms of ABM and their professional life. The majority of the respondents - 88%, wanted to apply ABM in the future in their research. Some of the respondents were already using ABM in their work. A large group (around 20-30%) was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities. Many respondents also shared the context to which they were thinking to apply ABM mentioning topics such as movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms. Others thought of demography, social networks, migration or settlement distributions patterns. The natural environment and the interaction with humans in the past was also mentioned frequently, often in relation to GIS and the question of how complimentary these two computational techniques are. The archaeological periods that the participants were interested in were very diverse, ranging from the Paleolithic to the Medieval period.

@@

As mentioned above, the following competence area's were identified as relevant for the tutorials: 1. Information and data literacy, 2. Communication and collaboration, 3. Digital content creation and, 5. Problem solving [@europeancommissionjointresearchcentre2022]. In these competence area's various competences are developed towards an advanced or even specialised level. A more extensive description of the Digital Competence Framework for Citizens in relation to Agent-Based Modelling for archaeologists is also available [@visser2024].

@@

Competence Area 3. Digital content creation is one of the main competence areas for the ABM-teaching material. In the process of learning ABM the users are constantly 3.1 Developing Digital Content and 3.2 Integrating And Re-Elaborating Digital Content. The material does not touch upon 3.3 Copyright And Licences, as they fall out of the scope of this project. A very important aspect of learning ABM is 3.4 Programming. The users are guidend through the process of developing the necessary programming skills to build a simulation. While the programming language used - NetLogo - is specific to ABM, its syntax follow general programming patterns and the learning is thus transferable to other programming languages. The users are expected to learn both syntax and more general computational concepts such as modular code development, testing and documenting code.

Competence Area 5. Problem solving is very important and is closely tied to 3.4 Programming, since programming involves a lot of problem solving but also the general principles of model development that have a prominent place in the educational materials. In addition, the creation of a highly complex ABM consists of problem solving all the time, including development of research questions, conceptualisation of theories, experiment design and validating against data - all of which require creative abilities and high level understanding of the scientific process. All competencies are addressed: 5.1 Solving Technical Problems, is common when writing code, learning about coding, error handling and going from pseudocode to real code. The competence 5.2 Identifying Needs And Technological Responses, is a central issue of the course, while 5.3 Creatively Using Digital Technology is an important aspect of so-called model thinking necessary to develop robust and insightful models. In tutorial 2 various aspects of 5.4 Identifying Digital Competence Gaps, are relevant, since this tutorial force the user to assess their knowledge and skills.

@@

The tutorials are publicly hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials/>) and are citeable [@rocks-macqueen2024]. Anyone interested can download the tutorials and their source code. To increase the accessibility and visibility of the tutorials, they are also hosted on a static website ([https://ABMArchaeologists.github.io/ABMA_Website](https://ABMArchaeologists.github.io/ABMA_Website/#/){.uri}) can be followed without the need to download any software. The source code of the website is also openly available [@jutte2024].

Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University [@scherjon2019] and external resources people can use to learn agent-based modelling. The website was mainly developed by a student group from Saxion University of Applied Sciences, see the acknowledgements.

@@

The process of creating and testing Open Educational Resources for Agent-Based Modelling for Archaeologists showed us many things. The demand for tutorials exceeded our expectations, showing the importance of creating these materials. The testing of tutorials during workshops at international an national conferences proved very useful in gaining direct feedback. This greatly helped improving the material. The tutorials have proven their use in guiding learners into understanding ABM in an archaeological context. The international cooperation improved the quality of the material and also the accessibility of the material for various audiences. The integration in the Digital competency framework and Skills passport increased the didactic and pedagogic quality even while the latter is not yet published.

@@

This project was undertaken with the support of the Erasmus+ programme of the European Union (<https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054>).

@@

Funding acknowledgements: the publication was financed by The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382); IR was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513, the Aarhus University Research Foundation and FWO grant number W001220N.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
# read data
pre_surveys <- readRDS("data/pre_workshop_surveys.rds")
post_surveys <- readRDS("data/post_workshop_surveys.rds")
# factore levels for charts
factor_levels <- c("Excellent", "Good", "Sufficient", "Insufficient", "Not shared")
# readable event names
event_names <- as_labeller(c("202304_CAA" = "CAA", "202308_EAA" = "EAA", "202310_CAA-DE-NLFl" = "CAA-DE-NLFl", "202311_Reuvensdagen" = "Reuvensdagen", "202311_CAA_UK" = "CAA_UK", "202312_Leiden" = "Leiden", "202401_Aarhus" = "Aarhus"))
events <- read.csv("data/events.csv", sep = ";")

@@

Computational modelling has become a standard research technique across most scientific disciplines, including archaeology. Among the many computational modelling techniques, the discipline has found agent-based modelling to be particularly well suited to the questions asked by archaeologists. Agent-based modelling is a simulation technique where autonomous agents interact with each other and with their environment. Their aggregated behaviour gives rise to system-level outcomes, thus enabling researchers to understand how *micro* (individual) *mechanisms* lead to *macro* (system-level) *patterns*.

While Agent-Based Modelling importance in archaeological practice is steadily growing [@romanowska2023] there are significant bariers to its wider adoption. One of the sectorial problems is the access to vocational training required to gain the necessary computer programming and modelling skills. A survey of practitioners [@davies2018] found significant deficiencies in the provision of ABM training; 76% of modellers were primarily self-taught with some degree of secondary support, i.e. 27% relied on peer support, only 7% had any sort of formal training at a workshop or summer school, while only 14% received training as part of a degree.

A consortium of Landward Research Teoranta, Aarhus Universitet, Leiden University and Saxion University of Applied Sciences recognised this need, and sought to address the limited opportunities that archaeologists have to access high quality training through a project aiming to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.

@@

-   development of a hands-on vocational training programme, aligned with the recently published textbook *Agent-Based Modelling for Archaeology: Simulating the Complexity of Societies* [@romanowska2021a];

@@

3.  improved opportunities for archaeologists to work in other European countries (transnational mobility) through matching the skills being delivered with the European Commission's Digital Competence Framework for Citizens (2.2) and the sectoral Skills Passport for Archaeology, which means that it will be easier for individual workers to demonstrate their skills to employers across Europe;

@@

The ABMA learning materials assume that participants have learning skills at least at a secondary education level. They are suitable for participants with at least some background in archaeology, such as those typical among students after the first year of a Bachelor study in Archaeology, i.e., above EQF levels 4 or 5. In addition, the tutorials are aimed at professionals working in archaeology who have some experience with computer applications. We also assume that learners are proficient in English and have at least the Reading B1 level, but Reading B2 is recommended [@councilofeurope2020]. The tutorials were also developed with a secondary aim to develop learners' general skills in the following competence areas: 1. Information and data literacy, 2. Communication and collaboration, 3. Digital content creation and, 5. Problem solving [@europeancommissionjointresearchcentre2022].

After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events all of which were run by members of the project team (Table \@ref(tab:events)). Recognising the difficulty of delivering practical training in an online environment during the online events several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things. Throughout the project we sought feedback from the participants and used it to improve the tutorials prior to the following events. The feedback was obtained through two surveys using Qualtrics (<https://www.qualtrics.com/>). The participants of the events were asked to respond to two surveys: one before the start of the event and one completed after the end of the event. The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix 1 for the questions). At the end of the event the participants were asked to answer a second survey. Questions of this survey were aimed at measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix 2 for the questions). For some events participants had to register beforehand and we were able to send the pre-workshop survey by email. At other events no registration was possible or necessary and the pre-workshop survey was shared at the start of the event. The post-workshop survey was distributed using QR-codes or links at the end of the event.

```{r, events, echo=FALSE, tab.cap= "Conferences, events and situations where workshops were held at which the tutorials were tested." }
knitr::kable(events, align = "l", col.names = c("Conference or event", "Period", "Online, in person or hybrid", "Number of participants", "Number of registrations")) 
```

@@

The final set of ABMA tutorials is available for download at <https://github.com/ABMArchaeologists/ABMA_tutorials> and is citeable [@rocks-macqueen2024]. The set consists of the following tutorials:

@@

Each tutorials consists of a number of individual lessons, each of which guides the learner in a self-paced manner through an increasingly challenging set of exercises that build the skills and knowledge necessary to become a confident agent-based modeller.

In the first tutorial the users learn what simulations and agent-based models are and how they can help in archaeological research (Figure \@ref(fig:screenshot-lesson-3)). This tutorial consists of four lessons and makes use of the Artificial Anasazi model [@axtell2002]. The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific. Various concepts related to ABM, such as the definitions of model and simulation, are introduced and the participants make their first steps in NetLogo syntax. The third lesson explains how ABMs are used in archaeological research. The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode. This tutorial builds a foundation for working in NetLogo and with ABM. All activity stays within the entry level of difficulty, although some higher level concepts are lightly touched upon.

@@

In tutorial 2 the user will learn the basics of NetLogo by making their simulation on the Out of Africal dispersal of *Homo sapiens* [@young1995].The learner works through the basics of NetLogo syntax and learns how to set up a simulation and visualize its outcomes. This tutorial consists of nine lessons. This tutorial intends to guide the learner from the introduction to the beginner level of proficiency. The learning curve is relatively gentle. This is achieved by working with the NetLogo web interface that enables highlighting specific words and buttons and giving visual cues as to how to proceed. The tutorial introduces several most commonly used built-in functions in Netlogo (called *primitives*) and introduces the general structure of all models consisting of the initialization phase and the main simulation loop. It walks the participants through the most important NetLogo settings, such as dimensions, coordinates and the origin point of the grid on which the simulation runs. The learners gain their first stripes in programming by using simple loops and functions as well as conditional statements. The next level is achieved by introducing the topic of function definition and different types of variables. Exporting data produced by the model through plots is also explained.

In the third tutorial, the learner builds a simple trade model, again based on a published simulation [@romanowska2018]. Here learners expand their NetLogo programming skills with more complexity added to such syntactic structures as loops, lists and reporters. They are introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity. This tutorial consists of seven lessons. It intends to guide the learner from the beginner to the intermediate level of proficiency. This is achieved by introducing standards of code development such as modular code, pseudocode, debugging and annotating the code. In addition, the tutorial introduces custom agent breeds, visualization with labels and reporters, plots and monitors. This takes the learner to a more advanced level, since it develops knowledge of the programming language combined with diagnosing unexpected behaviour and problem solving.

In tutorial 4 the learner works with the seminal Sugarscape model [@epstein1996] to further expand their agent-based modelling and NetLogo skills. The learner explores how to set up more complex interactions between agents and the environment. Furthermore, the principles of setting up good experiments and validating models is explained. This tutorial consists of eigth lessons. It intends to solidify the learner's understanding at the intermediate level of proficiency by introducing yet more complex agent-environment interaction, alternative ways to visualize the environment and modelling spatial dynamics. Creating toy landscapes is central to lesson 5. The learner dives into the topic of experiment design at a practical level but also with regards to the scientific principles, such as validation. Thus they are guided through the process of setting up experiments in NetLogo and collecting results using monitors and plots. The topic of validation of agent-based models is critical for most archaelogical applications and thus several ways to compare simulation results with the archaeological record are explored.

In the final and fifth tutorial the participant learns more about how to incorporate agent-based modelling in archaeological research. This tutorial focuses less on programming in NetLogo and more on the principle and standards of computer-based modelling. The learner learns about the different phases of the model development process. This tutorial consists of five lessons. It summarizes the material delivered in the previous lessons into one coherent framework consistent with an intermediate level of proficiency. This tutorial approaches more theoretical aspects in a practical environment. The learner revisits all phases of the model development process starting with the conceptual phase and finishing with the analysis and interpretation of results. There is a particular focus on discussing what kind of research questions are suitable for modelling and how to pick the right modelling technique. The importance of properly conceptualizing a model before starting the technical phase of the model development is highlighted as it is a common issue for less experienced modellers. The experiment design phase, including parametrisation, validation and the analysis and interpretation of the models is explained at length since it often proves challenging for students. On a practical side, the BehaviorSpace - NetLogo's experiment environment - is explained to enable the learners to batch run their models. Finally the dissemination phase of the model development is also included to ensure that the standards in model publication are widely known and the learners explore the fundamental importance of replication for advancing scientific knowledge.

@@

A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants). The respondents came from at least 40 countries that represented almost all continents (see Figure \@ref(fig:nationality)). There was a clear skew towards participants from Europe. The number of female respondents slightly outnumbered the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \@ref(fig:gender-age)). The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years old indicating that the audience was predominantly composed of early career archaeologists. It seems that the different events also had a slightly different distribution of both gender and age. For example, more older people attended the workshop at the CAA conference in April 2023 and more men were present during the workshop at the Reuvensdagen in November 2023. This might be due to the differences of audiences at the conferences. For the Reuvensdagen the number of participants of the workshop was relatively low and this variance might be due to chance.

@@

```{r gender-age, fig.cap="The gender and age distribution of the respondents for each workshop.", fig.width=11, echo=FALSE}
ggplot(pre_surveys, aes(x = `What.is.your.age.`, fill = `What.is.your.gender....Selected.Choice`)) +
  geom_bar() + labs(x= "Age", y="Count", fill = "Gender") + facet_grid(~Workshop, labeller = event_names) +   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

@@

```{r, available-theory, fig.cap="Respondents opinion on the quality of theory on ABM faceted out by event.", fig.width=11, echo=FALSE}
pre_surveys %>%
  select(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`, Workshop) %>%
  filter(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`!="") %>%
  mutate(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.` =
           fct_relevel(`How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`,
                       "No opinion", "Limited", "Sufficient", "More than sufficient", "Very good")) %>%
  ggplot(aes(x = `How.do.you.judge.the.available.theory.on.Agent.Based.Modelling.`)) +
  geom_bar() + xlab("Quality of theory on ABM") + coord_flip() +
  facet_grid(~Workshop, labeller = event_names)
```

@@

During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials (Figure \@ref(fig:workshop-impression)). Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research of discuss possibilities for the application of ABM in general. Individual progress varied depending on skills and preferences of the participants. While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later installments.

@@

The majority of the respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good. The teachers were even rated better than the teaching materials, with more respondents (165) giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).

@@

The respondents generally liked the interactive step-by-step way of going through the tutorials, but also valued the interaction with the teachers strongly. The easy and intuitive introduction in ABM and NetLOGO were also mentioned quite often. This was also observed during the workshops as most participants were working at their own pace and did not need much help. From a didactic point of view, the self-guided and self-paced way of going through the tutorials is very efficient for teachers and limits the frustration of students that may otherwise go too fast or fall behind the class when the tempo is dictated by the teacher.

We also received feedback on possible improvements. During the first workshop, the CAA conference in Amsterdam in April 2023, we did the first test of the tutorials and we also instructed the participants that this was work in progress. For this and the following few workshops we mostly received feedback about bugs causing the tutorial to crash or creating unexpected behaviour. The number of references to bugs reduced to zero by the end of the project and participants started providing other feedback, for example the wish to see more examples or expanding the tutorials further. Some wanted to introduce more cooperation between participants, which we did not consider primarily because the tutorials are aiming for individual work. A more cooperative aspect of the education materials is possible, even if harder for online events, and we hope that a wider classroom use of the ABMA materials will deliver solutions. It was also interesting to note that over time more and more people did not see any additional room for improvement indicating that the level, content and format of the tutorials met their needs.

We also asked the participants about their intentions in terms of ABM and their professional life. The majority of the respondents - 88%, wanted to apply ABM in the future in their research. Some of the respondents were already using ABM in their work. A large group (around 20-30%) was not sure yet how to apply ABM and wanted to read more on the subject or play a bit with the possibilities. Many respondents also shared the context to which they were thinking to apply ABM, mentioning topics such as movement of people or goods over land or water, sometimes in relation to trade or other distribution mechanisms. Others thought of demography, social networks, migration or settlement distribution patterns. The natural environment and the interaction with humans in the past was also mentioned frequently, often in relation to GIS and the question of how complimentary these two computational techniques are. The archaeological periods that the participants were interested in were very diverse, ranging from the Paleolithic to the Medieval period.

@@

As mentioned above, the following competence areas were identified as relevant for the tutorials: 1. Information and data literacy, 2. Communication and collaboration, 3. Digital content creation and, 5. Problem solving [@europeancommissionjointresearchcentre2022]. In these competence areas various competences are developed towards an advanced or even specialised level. A more extensive description of the Digital Competence Framework for Citizens in relation to Agent-Based Modelling for archaeologists is available elsewhere [@visser2024].

@@

Competence Area 3. Digital content creation is one of the main competence areas for the ABM teaching material. In the process of learning ABM the users are constantly 3.1 Developing Digital Content and 3.2 Integrating And Re-Elaborating Digital Content. The material does not touch upon 3.3 Copyright And Licences, as they fall out of the scope of this project. A very important aspect of learning ABM is 3.4 Programming. The users are guided through the process of developing the necessary programming skills to build a simulation. While the programming language used - NetLogo - is specific to ABM, its syntax follows general programming patterns and the learning is thus transferable to other programming languages. The users are expected to learn both syntax and more general computational concepts such as modular code development, testing and documenting code.

Competence Area 5. Problem solving is very important and is closely tied to 3.4 Programming, since programming involves a lot of problem solving but also the general principles of model development that have a prominent place in the educational materials. In addition, the creation of a highly complex ABM consists of problem solving all the time, including development of research questions, conceptualisation of theories, experiment design and validating against data - all of which require creative abilities and a high level understanding of the scientific process. All competencies are addressed: 5.1 Solving Technical Problems, is common when writing code, learning about coding, error handling and going from pseudocode to real code. The competence 5.2 Identifying Needs And Technological Responses, is a central issue of the course, while 5.3 Creatively Using Digital Technology is an important aspect of so-called model thinking necessary to develop robust and insightful models. In tutorial 2 various aspects of 5.4 Identifying Digital Competence Gaps, are relevant, since this tutorial forces the user to assess their knowledge and skills.

@@

The ABMA tutorials are publicly hosted on GitHub (<https://github.com/ABMArchaeologists/ABMA_tutorials/>) and are citeable [@rocks-macqueen2024]. Anyone interested can download the tutorials and their source code. To increase the accessibility and visibility of the tutorials, they are also hosted on a static website ([https://ABMArchaeologists.github.io/ABMA_Website](https://ABMArchaeologists.github.io/ABMA_Website/#/){.uri}) and can be followed without the need to download any software. The source code of the website is also openly available [@jutte2024].

Besides, the tutorial, the website provides: a link that can be used for feedback, basic explanations of agent-based modelling, an introductory presentation from one of the workshops, videos from an online course at Leiden University [@scherjon2019] and external resources that people can use to learn agent-based modelling. The website was mainly developed by a student group from Saxion University of Applied Sciences, see the acknowledgements.

@@

The process of creating and testing Open Educational Resources for Agent-Based Modelling for Archaeologists showed us many things. The demand for tutorials exceeded our expectations, showing the importance of creating these materials. The testing of tutorials during workshops at international and national conferences proved very useful in gaining direct feedback. This greatly helped with improving the materials. The tutorials have proven their use in guiding learners into understanding ABM in an archaeological context. The international cooperation improved the quality of the material and also the accessibility of the material for various audiences. The integration in the Digital competency framework and Skills passport increased the didactic and pedagogic quality even while the latter is not yet published.

@@

This project was undertaken with the support of the Erasmus+ programme of the European Union (<https://erasmus-plus.ec.europa.eu/nl/projects/search/details/2021-2-IE01-KA220-VET-000049054>). The publication was financed by The Carlsberg Foundation's *Young Researcher Fellowship* (CF21-0382); IR was supported by the European Union's Horizon 2020 Research and Innovation Programme under the Marie Skłodowska-Curie grant agreement no 754513, the Aarhus University Research Foundation and FWO grant number W001220N.

@@
